Inferring the existence and direction of causal associations in the face of measurement error

### 2017-01-03 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{31 December 2016}

@@

Measurement error will of course arise in more low-tech data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{12}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{8}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{13--15}.

Given the near-ubiquitous presence of measurement error in phenomic data
it is vital to understand its impact on the tools we use for causal
inference. An established study design that can provide information
about causality is randomisation. Given the hypothesis that trait A
(henceforth referred to as the exposure) is causally related to trait B
(henceforth referred to as the outcome), randomisation can be employed
to assess the causal nature of the association by randomly splitting the
sample into two groups, subjecting one group to the exposure and leaving
the other as a control. The association between the exposure and the
outcome in this setting provides a robust estimate of the causal
relationship. This provides the theoretical basis behind randomised
control trials, but in practice randomisation is often impossible to
implement in an experimental context due to cost, scale or inability to
manipulate the exposure. The principle, however, can be employed in
extant observational data through the use of genetic variants associated
with the exposure (instruments), where the inheritance of an allele
serves as a random lifetime allocation of differential exposure
levels\textsuperscript{3}. Two statistical approaches to exploiting the
properties of genetic instruments are widely used: mediation-based
approaches and Mendelian randomisation (MR).

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that the indirect influence of the SNP on the outcome will
disappear when conditioning on the exposure. Here, the exposure
`mediates' the association between the SNP and the outcome, providing
information about the causal influence of the exposure on the outcome.
This forms the basis of a number of methods such as genetical
genomics\textsuperscript{16}, the regression-based causal inference test
(CIT)\textsuperscript{4,17}, a structural equation modelling (SEM)
implementation in the NEO software\textsuperscript{5}, and various other
methods including Bayesian approaches\textsuperscript{6}. They have been
employed by a number of recent publications that make causal inferences
in large scale `omics datasets\textsuperscript{6,18--22}.

@@

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. The main advantage of
mediation over MR is that the true underlying biology of the genetic
instrument doesn't necessarily need to be known as long as it associates
with the putative exposure. In the CIT framework (described fully in the
Methods) for example, the test statistic is different if you test for
the exposure causing the outcome or the outcome causing the exposure,
allowing the researcher to infer the direction of causality between two
variables by performing the test in both directions and choosing the
model with the strongest evidence.

Such is not the case for MR, where in order to infer the direction of
causality the instrument must be known to influence the exposure
directly, associating with the outcome only through the exposure. This
requirement can be problematic because if there exists a putative
association between two variables, with the instrumental SNP being
robustly associated with each, it can be difficult to determine which of
the two variables is subject to the direct effect of the SNP (i.e., for
which of the two variables is the SNP a valid instrument? Figure 1). By
definition, we expect that if the association is causal then a SNP for
the exposure will be associated with the outcome, so if the researcher
erroneously uses the SNP as an instrument for the outcome then they are
likely to see an apparently robust causal association of outcome on
exposure. Such a situation can arise in many scenarios. Genome-wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit\textsuperscript{23}. For
example, the same FTO variant that is associated with body mass index
(BMI) is also associated with C-reactive protein (CRP). It is known that
this is because BMI causes CRP\textsuperscript{24}, but supposing we
only had \emph{a priori} knowledge that FTO associated with CRP, and we
used it to instrument CRP in an MR analysis of CRP against BMI, we would
mistakenly infer that CRP causes BMI. Another situation is where the
causal directions between 'omic measures need to be determined, for
example if a DNA methylation probe is associated with expression of an
adjacent gene, then is a cis-acting SNP an instrument for the DNA
methylation level, or the gene expression level\textsuperscript{25}
(Figure 1)?

MR does however have some important advantages over mediation based
approaches. First, MR is used as a tool to not just deduce the existance
of causality, but to also estimate the causal effect size. Second, using
mediation requires that the exposure, outcome and instrument variables
are all measured in the same data, whereas recent extensions to MR
circumvent this requirement, allowing causal inference to be drawn when
exposure variables and outcome variables are measured in different
samples\textsuperscript{26}. This has the crucial advantage of improving
statistical power by allowing analysis in much larger sample sizes, and
dramatically expands the breadth of possible phenotypic relationships
that can be evaluated. Third, MR is substantially more robust to there
being measurement error in the exposure and outcome variables. Indeed
instrumental variable (IV) analysis was in part initially introduced as
a correction for measurement error\textsuperscript{27}, whereas it has
been noted that both classic mediation-based
analyses\textsuperscript{14,15,28,29} and mediation-based methods that
use instrumental variables\textsuperscript{30,31} are prone to be
unreliable in its presence. We argue that this is a serious limitation.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{32} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{4} is implemented in the R package
\emph{R/cit}\textsuperscript{17}. The methodology of the CIT is as
follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
the exposure \(x\) is causally related to an outcome \(y\). Thus,

@@

where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
residual of \(y\) after adjusting for the \(x\), where \(x\) is assumed
to mediate the association between the SNP and the outcome. If all four
tests reject the null hypothesis then it is inferred that \(x\) is
causally related to \(y\). The CIT measures the strength of causality by
generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
(least extreme) p-value of the four tests, the intuition being that
causal inference is only as strong as the weakest link in the chain of
tests.

In these analyses the \emph{cit.cp} function was used to obtain an
omnibus p-value. To infer the direction of causality using the CIT
method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
for the correct model of \(x\) causing \(y\), and for incorrect model of
\(y\) causing \(x\). For some significance threshold \(\alpha\), the
existence of causality and its direction was inferred based on the
following scenarios:

@@

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we should stress that for real analyses
it is not good practice to use p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds.

@@

where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
R package \(R/systemfit\)\textsuperscript{33}.

The direction of the causal association was inferred as follows.
Assuming no horizontal pleiotropy (i.e. \(g\) only has an effect on one
of the variables through the other variable) it is desirable to know
which of the variables, \(x\) or \(y\), it has a direct influence on.
This can be achieved by assessing which of the two variables has the
biggest absolute correlation with \(g\) (Appendix 2). This test can be
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population\textsuperscript{34}. It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{26} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (or from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package
\emph{R/psych}\textsuperscript{35}.

@@

Simulations were conducting by creating variables of sample size \(n\)
for the exposure \(x\), the measured values of the exposure \(x_o\), the
outcome \(y\), the measured values of the outcome \(y_o\) and the
instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
instrument for \(x\). Each variable was simulated such that:

@@

Two sample MR\textsuperscript{26} was performed using the summary
statistics presented in\textsuperscript{25} for each of 4122
associations between DNA methylation and gene expression levels. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

@@

\subsubsection{MR analysis of gene expression and DNA
methylation}\label{mr-analysis-of-gene-expression-and-dna-methylation}

We used the Steiger test to infer the direction of causality between DNA
methylation levels and gene expression levels. To do this we obtained a
list of 458 gene expression - DNA methylation associations as reported
in Shakhbazov et al\textsuperscript{25}. These were filtered to be
located on the same chromosome, have robost correlations after
correcting for multiple testing, and to share a SNP that had a robust
cis-acting effect on both the DNA methylation probe and the gene
expression probe. Using the reported summary data we used the Steiger
test of two independent correlations to identify which direction of
causality was most likely for each association. We found that the causal
direction commonly goes in both directions (Table 1) but methylation
levels were more likely to influence expression levels than vice versa
(probability of expression causing methylation = 0.0163873 (95\% CI
0.3738727 - 0.4874269) amongst NA associations with
\(p_{Steiger} < 0.05\)).

We performed two sample MR\textsuperscript{26} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
always in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{25}, and that there was a
moderate correlation between the magnitude of the causal correlation and
the observational Pearson correlation (Table 1).

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.5326663
(95\% CI 0.3172203 - 0.8851868)), suggesting that reducing gene
expression levels at a controlling CpG typically leads to increased gene
expression levels, consistent with expectation\textsuperscript{36}.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data, and
using genetic instruments as surrogates for randomisation is a useful
approach. MR and the CIT employ genetic instruments in different ways,
and unfortunately neither offer solutions that work perfectly in all
scenarios. In the face of measurement error it is evident that causal
inference drawn through mediation based methods such as the CIT analysis
typically fare worse than MR and will be difficult to interpret, and
this is a serious problem considering that it is often impossible to
estimate the extent of measurement error present in observational data.

In the epidemiological literature issues of measurement error in
mediation is relatively well explored\textsuperscript{13}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though seemingly
unintuitive, increasing sample size does not resolve the issue, rather
it leads to more extreme p-values for the model that predicts the wrong
direction of causality.

@@

Mediation based network approaches are very well
established\textsuperscript{31} and have a number of extensions that
make them valuable tools, including for example network construction.
But because they are predicated on the basic underlying principles of
mediation they are liable to suffer from the same issues of measurement
error. Recent advances in MR methodology, for example applying MR to
genetical genomics\textsuperscript{37}, multivariate
MR\textsuperscript{38}, network MR\textsuperscript{39} and mediation
through MR\textsuperscript{40} may offer more robust alternatives for
these more complicated problems.

In our simulations we focused on the simple case of a single instrument
in a single sample setting, and assumed that pleiotropy (the influence
of the instrument on the outcome through a mechanism other than the
exposure) was not present. Recent method developments in
MR\textsuperscript{41} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

Our analysis of DNA methylation and gene expression summary data
demonstates the simplicity of inferring causal existence, direction and
magnitude in the two sample MR framework. We note that verifying these
results with regards to the extent to which measurement error has
influenced them empirically is a challenging task and beyond the scope
of this work.

@@

\begin{verbatim}
## $r2_exp
## [1] 0.1
## 
## $r2_out
## [1] 0.03162278
## 
## $r2_exp_adj
## [1] 0.1
## 
## $r2_out_adj
## [1] 0.03162278
## 
## $correct_causal_direction
## [1] TRUE
## 
## $steiger_test
## [1] 0
## 
## $correct_causal_direction_adj
## [1] TRUE
## 
## $steiger_test_adj
## [1] 0
## 
## $vz
## [1] -0.08934366
## 
## $vz0
## [1] 0.0262863
## 
## $vz1
## [1] 0.11563
## 
## $sensitivity
## [1] 0.185224
## 
## $sensitivity_ratio
## [1] 4.398868
## 
## $sensitivity_plot
\end{verbatim}

\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-1.pdf}
\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-2.pdf}
\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-3.pdf}

@@

Figure 6: Using 458 putative associations between DNA methylation and
gene expression we used the Steiger test to infer the direction of
causality between them. The rightmost bar shows the proportion of
associations for each of the two possible causal directions (colour key)
assuming no measurement error in either gene expression or DNA
methylation levels. The proportions change when we assume different
levels of measurement error in gene expression levels (x-axis) or DNA
methylation levels (columns of boxes). If there is systematically higher
measurement error in one platform it will appear to be less likely to be
the causal factor.

@@

In the presence of measurement error in \(x\) and \(y\), however, the
Steiger test will instead be assessing the inequality
\(\rho_{g, x_o} > \rho_{g, y_o}\). In order to assess how reliable the
Steiger test is in the presence of measurement imprecision, we can
evaluate the range of potential values of measurement error in \(x\) and
\(y\) over which the empirical result from the Steiger test would return
the wrong value.

@@

Where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error range in
\(x_o\) and \(y_o\) will not change the qualitative inference of the
direction of causality. To evaluate the reliability, \(R\), of the
Steiger test to measurement error, the objective is to compare the
weighted proportion of the parameter space that agrees with the inferred
direction against the weighted proportion which does not:

@@

\hypertarget{ref-Fewell2007}{}
13. Fewell, Z., Davey Smith, G. \& Sterne, J. A. C. The impact of
residual and unmeasured confounding in epidemiologic studies: a
simulation study. \emph{American journal of epidemiology} \textbf{166,}
646--55 (2007).

\hypertarget{ref-LeCessie2012}{}
14. Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
Vandenbroucke, J. P. Quantification of bias in direct effects estimates
due to different types of measurement error in the mediator.
\emph{Epidemiology (Cambridge, Mass.)} \textbf{23,} 551--60 (2012).

\hypertarget{ref-Blakely2013}{}
15. Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
mediator matters when estimating indirect effects. \emph{Journal of
epidemiology and community health} \textbf{67,} 458--66 (2013).

\hypertarget{ref-Schadt2005}{}
16. Schadt, E. E. \emph{et al.} An integrative genomics approach to
infer causal associations between gene expression and disease.
\emph{Nature Genetics} \textbf{37,} 710--717 (2005).

\hypertarget{ref-Millstein2016}{}
17. Millstein, J. cit: Causal Inference Test. R package version 1.9.
(2016). at
\textless{}\url{http://cran.r-project.org/package=cit}\textgreater{}

>>>>

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{03 January 2017}

@@

Measurement error will of course arise in more basic data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{12}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{8}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{13,14}.

Given the near-ubiquitous presence of measurement error in phenomic data
it is vital to understand its impact on the tools we use for causal
inference. An established study design that can provide information
about causality is randomisation. Given the hypothesis that trait A
(henceforth referred to as the exposure) is causally related to trait B
(henceforth referred to as the outcome), randomisation can be employed
to assess the causal nature of the association by randomly splitting the
sample into two groups, subjecting one group to the exposure and leaving
the other as a control. The association between the exposure and the
outcome in this setting provides a robust estimate of the causal
relationship. This provides the theoretical basis behind randomised
control trials, but in practice randomisation is often impossible to
implement in an experimental context due to cost, scale or inability to
manipulate the exposure. The principle, however, can be employed in
extant observational data through the use of genetic variants associated
with the exposure (instruments), where the inheritance of an allele
serves as a random lifetime allocation of differential exposure
levels\textsuperscript{15,16}. Two statistical approaches to exploiting
the properties of genetic instruments are widely used: mediation-based
approaches and Mendelian randomisation (MR).

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that the indirect influence of the SNP on the outcome will
disappear when conditioning on the exposure. Here, the exposure
`mediates' the association between the SNP and the outcome, providing
information about the causal influence of the exposure on the outcome.
This forms the basis of a number of methods such as genetical
genomics\textsuperscript{17}, the regression-based causal inference test
(CIT)\textsuperscript{4,18}, a structural equation modelling (SEM)
implementation in the NEO software\textsuperscript{5}, and various other
methods including Bayesian approaches\textsuperscript{6}. They have been
employed by a number of recent publications that make causal inferences
in large scale `omics datasets\textsuperscript{6,19--23}.

@@

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. An important advantage
of mediation-based analysis is that the true underlying biology of the
genetic instrument doesn't necessarily need to be known as long as it
associates with the putative exposure. In the CIT framework (described
fully in the Methods) for example, the test statistic is different if
you test for the exposure causing the outcome or the outcome causing the
exposure, allowing the researcher to infer the direction of causality
between two variables by performing the test in both directions and
choosing the model with the strongest evidence. Such is not the case for
MR, where in order to infer the direction of causality the instrument
must be known to influence the exposure directly, associating with the
outcome only through the exposure.

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the instrumental SNP being robustly associated with each, it can be
difficult to determine which of the two variables is subject to the
direct effect of the SNP (i.e.~for which of the two variables is the SNP
a valid instrument? Figure 1). By definition, we expect that if the
association is causal then a SNP for the exposure will be associated
with the outcome, so if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. Genome-wide association studies (GWASs) that
identify genetic associations for complex traits are, by design,
hypothesis free and agnostic of genomic function, and it often takes
years of follow up studies to understand the biological nature of a
putative GWAS hit\textsuperscript{24}. For example, the same FTO variant
that is associated with body mass index (BMI) is also associated with
C-reactive protein (CRP). It is known that this is because BMI causes
CRP\textsuperscript{25}, but supposing we only had \emph{a priori}
knowledge that FTO associated with CRP, and we used it to instrument CRP
in an MR analysis of CRP against BMI, we would mistakenly infer that CRP
causes BMI. If multiple instruments are available for an hypothesised
exposure then this problem can be ameliorated\textsuperscript{16},
however this is not always the case, especially in the case of
determining causal directions between 'omic measures where typically
only one cis-acting SNP is known. For example if a DNA methylation probe
is associated with expression of an adjacent gene, then is a cis-acting
SNP an instrument for the DNA methylation level, or the gene expression
level\textsuperscript{26} (Figure 1)?

MR does however have some important advantages over the mediation-based
approaches. First, MR is used as a tool to not just deduce the existance
of causality, but to also estimate the causal effect size. Second, the
mediation-based approaches require that the exposure, outcome and
instrumental variables are all measured in the same data, whereas recent
extensions to MR circumvent this requirement, allowing causal inference
to be drawn when exposure variables and outcome variables are measured
in different samples\textsuperscript{27}. This has the crucial advantage
of improving statistical power by allowing analysis in much larger
sample sizes, and dramatically expands the breadth of possible
phenotypic relationships that can be evaluated\textsuperscript{28}.
Third, MR is substantially more robust to there being measurement error
in the exposure variable. Indeed instrumental variable (IV) analysis was
in part initially introduced as a correction for measurement error in
the exposure\textsuperscript{29}, whereas it has been noted that both
classic mediation-based analyses\textsuperscript{13,14,30,31} and
mediation-based methods that use instrumental
variables\textsuperscript{32,33} are prone to be unreliable in its
presence.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{34} and code is made
available at \href{}{github url to go here} and implemented in the
MR-Base (\url{http://wwww.mrbase.org}) platform\textsuperscript{28}.

@@

The CIT method\textsuperscript{4} is implemented in the R package
\emph{R/cit}\textsuperscript{18}. The methodology of the CIT is as
follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
the exposure \(x\) is causally related to an outcome \(y\). Thus,

@@

where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
mediate the association between the SNP and the outcome. If all four
tests reject the null hypothesis then it is inferred that \(x\) is
causally related to \(y\). The CIT measures the strength of causality by
generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
(least extreme) p-value of the four tests, the intuition being that
causal inference is only as strong as the weakest link in the chain of
tests.

In these analyses the \emph{cit.cp} function was used to obtain an
omnibus p-value. To infer the direction of causality using the CIT
method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
for the correct model of \(x\) causing \(y\), and for the incorrect
model of \(y\) causing \(x\). For some significance threshold
\(\alpha\), the existence of causality and its direction was inferred
based on the following scenarios:

@@

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we should stress that for real analyses
it is not good practice to rely on p-values for making causal inference,
nor is it reliable to depend on arbitrary significance
thresholds\textsuperscript{35}.

@@

where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
R package \(R/systemfit\)\textsuperscript{36}.

In what follows we assume no horizontal pleiotropy, such that \(g\) only
has an effect on one of the variables through the other variable. To
infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Appendix 2),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population\textsuperscript{37}. It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{27} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (e.g.~from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package
\emph{R/psych}\textsuperscript{38}.

@@

Simulations were conducted by creating variables of sample size \(n\)
for the exposure \(x\), the measured values of the exposure \(x_o\), the
outcome \(y\), the measured values of the outcome \(y_o\) and the
instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
instrument for \(x\). Each variable was simulated such that:

@@

Two sample MR\textsuperscript{27} was performed using the summary
statistics presented in\textsuperscript{26} for each of 4122
associations between DNA methylation and gene expression levels. To do
this we obtained a list of 458 gene expression - DNA methylation
associations as reported in Shakhbazov et al\textsuperscript{26}. These
were filtered to be located on the same chromosome, have robost
correlations after correcting for multiple testing, and to share a SNP
that had a robust cis-acting effect on both the DNA methylation probe
and the gene expression probe. Because only summary statistics were
available (effect, standard error, effect allele, sample size, p-values)
for the instrumental SNP on the methylation and gene expression levels,
the Steiger test of two independent correlations was used to infer the
direction of causality for each of the associations. The Wald ratio test
was then used to estimate the causal effect size for the estimated
direction for each association.

@@

\subsubsection{MR analysis of gene expression and DNA methylation
levels}\label{mr-analysis-of-gene-expression-and-dna-methylation-levels}

We used the Steiger test to infer the direction of causality between DNA
methylation levels and gene expression levels and found that the causal
direction commonly goes in both directions (Figure 6a). We then went on
to predict the causal directions of the associations for varying levels
of possible measurement error. Figure 6a shows that the qualitative
result of which of DNA methylation or gene expression is more commonly
the causal factor is very strongly determined by measurement error.

We performed two sample MR\textsuperscript{27} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
generally in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{26}, but with a few
exceptions (Figure 6b). There was a moderate correlation between the
absolute magnitudes of the causal correlation and the observational
Pearson correlation (r = 0.4423627). Together these inferences suggest
that even for relatively low level associations it is important to use
causal inference methods over observational associations to infer causal
effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.5326663
(95\% CI 0.3172203 - 0.8851868), Figure 6c), suggesting that reducing
gene expression levels at a controlling CpG typically leads to increased
gene expression levels, consistent with expectation\textsuperscript{39}.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored\textsuperscript{40}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this
position\textsuperscript{41,42}, though making strong conclusions for
this analysis is difficult because measurement error is likely to be
study specific.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
MR strategies exist for other scenarios, for example if there are
multiple instruments for the hypothesised exposure then the problems
described here may be circumvented\textsuperscript{16}. Likewise, if
there is at least one instrument for each trait then bi-directional MR
can offer solutions to inferring the causal
direction\textsuperscript{43}. In this work we assumed that pleiotropy
(the influence of the instrument on the outcome through a mechanism
other than the exposure) was not present. Recent method developments in
MR\textsuperscript{44,45} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established\textsuperscript{33} and have a
number of extensions that make them valuable tools, including for
example network construction. But because they are predicated on the
basic underlying principles of mediation they are liable to suffer from
the same issues of measurement error. Recent advances in MR methodology,
for example applying MR to genetical genomics\textsuperscript{46},
multivariate MR\textsuperscript{47} and mediation through
MR\textsuperscript{48,49} may offer more robust alternatives for these
more complicated problems.

@@

\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-1.pdf}

@@

Figure 6: Using 458 putative associations between DNA methylation and
gene expression we used the Steiger test to infer the direction of
causality between them. a) The rightmost bar shows the proportion of
associations for each of the two possible causal directions (colour key)
assuming no measurement error in either gene expression or DNA
methylation levels. The proportions change when we assume different
levels of measurement error in gene expression levels (x-axis) or DNA
methylation levels (columns of boxes). If there is systematically higher
measurement error in one platform it will appear to be less likely to be
the causal factor. b) The relationship between the Pearson correlation
between DNA methylation and gene expression levels (x-axis) and the
causal estimate (scaled to be in standard deviation units, y-axis). c)
Distribution of estimated causal effect sizes, stratified into
associations inferred to be due to DNA methylation causing expression
(blue) and expression causing DNA methylation (red).

@@

In the presence of measurement error in \(x\) and \(y\), however, the
Steiger test will instead be assessing the inequality
\(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:

\[
\begin{aligned}
\rho_{g, x_O} & > \rho_{g, y_O} \\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
\end{aligned}
\]

In order to assess how reliable the Steiger test is in the presence of
measurement imprecision, we can evaluate the range of potential values
of measurement error in \(x\) and \(y\) over which the empirical result
from the Steiger test would return the wrong value.

@@

Where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error range in
\(x_o\) and \(y_o\) will not change the qualitative inference of the
direction of causality. To evaluate the reliability, \(R\), of the
Steiger test with regards to measurement error, the objective is to
compare the weighted proportion of the parameter space that agrees with
the inferred direction against the weighted proportion which does not:

@@

\hypertarget{ref-LeCessie2012}{}
13. Cessie, S. le, Debeij, J., Rosendaal, F. R., Cannegieter, S. C. \&
Vandenbroucke, J. P. Quantification of bias in direct effects estimates
due to different types of measurement error in the mediator.
\emph{Epidemiology (Cambridge, Mass.)} \textbf{23,} 551--60 (2012).

\hypertarget{ref-Blakely2013}{}
14. Blakely, T., McKenzie, S. \& Carter, K. Misclassification of the
mediator matters when estimating indirect effects. \emph{Journal of
epidemiology and community health} \textbf{67,} 458--66 (2013).

\hypertarget{ref-DaveySmith2003}{}
15. Davey Smith, G. \& Ebrahim, S. 'Mendelian randomization': can
genetic epidemiology contribute to understanding environmental
determinants of disease? \emph{International Journal of Epidemiology}
\textbf{32,} 1--22 (2003).

\hypertarget{ref-DaveySmithHemani2014}{}
16. Davey Smith, G. \& Hemani, G. Mendelian randomization: genetic
anchors for causal inference in epidemiological studies. \emph{Human
molecular genetics} \textbf{23,} R89-----R98 (2014).

\hypertarget{ref-Schadt2005}{}
17. Schadt, E. E. \emph{et al.} An integrative genomics approach to
infer causal associations between gene expression and disease.
\emph{Nature Genetics} \textbf{37,} 710--717 (2005).
